# Advanced {#advanced}

The following chapter provides advanced use cases of the `tidyvpc` package.

```{r init, include=FALSE}
library(tidyvpc)
obs_data <- as.data.table(obs_data)
sim_data <- as.data.table(sim_data)
obs_data <- obs_data[obs_data$MDV == 0,]
sim_data <- sim_data[sim_data$MDV == 0,]
obs_data$PRED <- sim_data$PRED[sim_data$REP == 1]

```

## Binning by Strata

To use different binning methods for different stratification variables, and/or for each level of stratification variable, use multiple calls to the `binning()` function in combination with the `stratum` argument. Make sure to set `by.strata = T`

```{r bin-strata, fig.width = 9, fig.height = 6, out.width=640}
vpc <- observed(obs_data, x=TIME, y=DV) %>%
    simulated(sim_data, y=DV) %>%
    stratify(~ GENDER + STUDY) %>%
    binning(stratum = list(GENDER = "M", STUDY = "Study A"), bin = "jenks", nbins = 5, by.strata = T) %>%
    binning(stratum = list(GENDER = "F", STUDY = "Study A"), bin = "centers", centers = c(0.5,3,5,10,15), by.strata = T) %>%
    binning(stratum = list(GENDER = "M", STUDY = "Study B"), bin = "kmeans", by.strata = T) %>%
    binning(stratum = list(GENDER = "F", STUDY = "Study B"), bin = "pam", nbins = 5, by.strata = T) %>%
    predcorrect(pred=PRED) %>%
    vpcstats()

plot(vpc)
```

## Binless by Strata

To use different smoothing parameters for each level of stratification variable if using the `binless()` function use a single call of the `binless()` function and include a `data.frame` with the column names of stratification variable and corresponding level. To use different `span` values for each level of stratification variable use a vector the length of n levels of strata. Note: If using more than one stratification variable with the `binless()` function you must set `optimize = TRUE` and optimize lambda and span using AIC.

```{r binless-strata, warning = FALSE, fig.width = 9, fig.height = 6, out.width=640}
new_lambda = data.frame(GENDER_F = c(2,4,2), GENDER_M = c(1.9,3,2.25) )

vpc <- observed(obs_data, x=TIME, y=DV) %>%
    simulated(sim_data, y=DV) %>%
    stratify(~ GENDER) %>%
    predcorrect(pred=PRED) %>%
    binless(qpred = c(0.1, 0.5, 0.9), optimize = FALSE, lambda = new_lambda, loess.ypc = TRUE, span = c(.6, .85)) %>%
    vpcstats()

plot(vpc)
```


## Visualize Bins

If using `binning()` methods, you can visualize bins by using the `plot()` function on the `tidyvpcobj` without calling `vpcstats()`. Once you are satisifed with the binning method, simply call `vpcstats()` on the existing `tidyvpcobj` to compute VPC percentiles and prediction intervals i.e `vpc %>% vpcstats()`

```{r bin-visualize, warning = FALSE, fig.width = 9, fig.height = 6, out.width=640}
vpc <- observed(obs_data, x=TIME, y=DV) %>%
    simulated(sim_data, y=DV) %>%
    binning(bin = "jenks", nbins = 7)

plot(vpc)

```

## Bin Information

To obtain information about the bins, including the number of observations, xmedian, xmean, xmin, xmax, xmidpoint, xleft, xright, and xcenter, use the `bininfo()` function from `tidyvpc`.

```{r bin-info}
vpc <- observed(obs_data, x=TIME, y=DV) %>%
    simulated(sim_data, y=DV) %>%
    binning(bin = "jenks", nbins = 4) %>%
    vpcstats()

bin_information <- bininfo(vpc)
head(bin_information)

```
