# Extending Further with `ggplot2` {#ggplot2}

```{r include=FALSE}
library(tidyvpc)
library(ggplot2)
library(data.table, quietly = TRUE, warn.conflicts = FALSE)

assignInNamespace("cedta.pkgEvalsUserCode", c(data.table:::cedta.pkgEvalsUserCode, "rtvs"), "data.table")

obs_data <- as.data.table(obs_data)
sim_data <- as.data.table(sim_data)
obs_data <- obs_data[obs_data$MDV == 0,]
sim_data <- sim_data[sim_data$MDV == 0,]
obs_data$PRED <- sim_data$PRED[sim_data$REP == 1]
```

While the built in `plot()` function make it easy to quickly visualize the derived VPC, the `tidyvpcobj` can be plotted using `ggplot2` for complete plot customization.

## Plot VPC

```{r fig.width = 9, fig.height = 6, out.width=640, warning=FALSE}
obs_data$LLOQ <- ifelse(obs_data$STUDY == "Study A", 50, 25)

vpc <- observed(obs_data, x = TIME, y = DV) %>%
  simulated(sim_data, y = DV) %>%
  censoring(blq = DV < LLOQ, lloq = LLOQ) %>%
  stratify(~STUDY) %>%
  binning(bin = NTIME) %>%
  vpcstats(qpred = c(0.1, 0.5, 0.9))

ggplot(vpc$stats, aes(x = xbin)) + 
  facet_grid(~STUDY, scales = "free", as.table = FALSE) + 
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = qname, col = qname, group = qname),alpha = 0.1, col = NA) + 
  geom_line(aes(y = md, col = qname, group = qname)) +
  geom_line(aes(y = y, linetype = qname), size = 1) + 
  geom_hline(data=unique(obs_data[, .(STUDY, LLOQ)]), aes(yintercept=LLOQ), linetype="dotted", size=1) +
  geom_text(data = unique(vpc$data[, .(LLOQ), by = "STUDY"]), 
            aes(x = 10, y = LLOQ, label = paste("LLOQ", LLOQ, sep = "="), ), vjust = 1, hjust = 1) +
  scale_colour_manual(name = "Simulated Percentiles\nMedian (lines) 95% CI (areas)",
                    breaks = c("q0.1", "q0.5", "q0.9"), 
                    values = c("red", "blue", "red"), 
                    labels = c("10%", "50%", "90%")) + 
  scale_fill_manual(name = "Simulated Percentiles\nMedian (lines) 95% CI (areas)", 
                    breaks = c("q0.1", "q0.5", "q0.9"), 
                    values = c("red", "blue", "red"), 
                    labels = c("10%", "50%", "90%")) + 
  scale_linetype_manual(name = "Observed Percentiles\nMedian (lines) 95% CI (areas)", 
                    breaks = c("q0.1", "q0.5", "q0.9"), 
                    values = c("dotted", "solid", "dashed"), 
                    labels = c("10%", "50%", "90%")) + 
  guides(fill = guide_legend(order = 2), colour = guide_legend(order = 2), linetype = guide_legend(order = 1)) + 
  theme(legend.position = "top", legend.key.width = grid::unit(1, "cm")) + 
  labs(x = "TIME", y = "Concentration") + 
  geom_point(data = vpc$obs, aes(x = x, y = y), size = 1, alpha = 0.1, show.legend = FALSE) + 
  geom_vline(data = bininfo(vpc)[, .(x = sort(unique(c(xleft, xright)))), by = names(vpc$strat)],aes(xintercept = x), size = rel(0.5), col = "gray80") + 
  theme(panel.grid = element_blank()) + 
  geom_rug(data = bininfo(vpc)[, .(x = sort(unique(c(xleft, xright)))), by = names(vpc$strat)],aes(x = x), sides = "t", size = 1)


```


## Plot Rectangles

The results from `bininfo()` make it easy to plot a rectangle VPC using `ggplot2`.

```{r warning=FALSE}
vpc <- observed(obs_data, x=TIME, y=DV) %>%
    simulated(sim_data, y=DV) %>%
    binning(bin = "jenks", nbins = 4) %>%
    vpcstats()

#Get vpcstats df
stats <- vpc$stats
#Get bininfo df
bin_information <- bininfo(vpc)
#Left join bin_info to vpcstats on bin
bin_information <- stats[bin_information, on = "bin"]
#Generate ymin
bin_information <- bin_information[, ymin := min(y), by = "bin"]
#Generate ymax
bin_information <- bin_information[, ymax := max(y), by = "bin"]
head(bin_information)
```


Plot rectangles using `ymin` and `ymax`, the min/max y values in `vpc$stats` grouped by bin.
```{r fig.width = 9, fig.height = 6, out.width=640, warning=FALSE}
ggplot(bin_information, aes(x = xbin)) + 
  geom_line(aes(y = md, col = qname, group = qname)) +
  geom_line(aes(y = y, linetype = qname), size = 1) +
  geom_rect(aes(xmin= xleft,xmax= xright, ymin =  ymin, ymax =  ymax),alpha = .1, col = "black", fill = "green") +
  geom_point(data = vpc$obs, aes(x = x, y = y), size = 1, alpha = 0.1, show.legend = FALSE) + 
  scale_colour_manual(name = "Simulated Percentiles\nMedian (lines) 95% CI (areas)",
                    breaks = c("q0.05", "q0.5", "q0.95"), 
                    values = c("red", "blue", "red"), 
                    labels = c("5%", "50%", "95%")) + 
  scale_linetype_manual(name = "Observed Percentiles\nMedian (lines) 95% CI (areas)", 
                    breaks = c("q0.05", "q0.5", "q0.95"), 
                    values = c("dotted", "solid", "dashed"), 
                    labels = c("5%", "50%", "95%")) + 
  geom_vline(data = bin_information[, list(x = sort(unique(c(xleft, xright))))],aes(xintercept = x), size = rel(0.5), col = "gray80") +
  geom_rug(data = bin_information[, list(x = sort(unique(c(xleft, xright))))],aes(x = x), sides = "t", size = 1) +
  guides(fill = guide_legend(order = 2), colour = guide_legend(order = 2), linetype = guide_legend(order = 1)) + 
  theme(legend.position = "top", legend.key.width = grid::unit(1, "cm")) + 
  labs(x = "TIME", y = "Concentration")

      
```

Alternatively, we can obtain the required data for plotting used in the above `bin_information` data frame by merging `vpc$stats` and `bininfo(vpc)` on `bin` in the `ggplot2` data argument.  If stratifying you will need to include the name of the stratification variable(s) in the `data.table` merge i.e. `vpc$stats[bininfo(vpc), on=c("STUDY", "bin")]`. In the rectangle vpc below, we will stratify on `STUDY` and plot rectangles for each quantile.

```{r fig.width = 9, fig.height = 6, out.width=640, warning=FALSE}
obs_data$LLOQ <- obs_data[, ifelse(STUDY == "Study A", 50, 25)]

vpc <- observed(obs_data, x = TIME, y = DV) %>%
  simulated(sim_data, y = DV) %>%
  censoring(blq = DV < LLOQ, lloq = LLOQ) %>%
  stratify(~STUDY) %>%
  binning(bin = NTIME) %>%
  vpcstats(qpred = c(0.1, 0.5, 0.9))


ggplot(vpc$stats[bininfo(vpc), on=c("STUDY", "bin")], aes(x = xbin)) + 
  facet_grid(~STUDY, scales = "free", as.table = FALSE) + 
  geom_rect(aes(xmin = xleft, xmax = xright, ymin = lo, ymax = hi, fill = qname, col = qname, group = qname),alpha = 0.1, col = NA) + 
  geom_segment(aes(x = xleft, xend = xright, y = md, yend = md, col = qname, group = qname)) +
  geom_segment(aes(x = xleft, xend = xright, y = y, yend = y, linetype = qname), size = 1) +
  geom_line(aes(y = md, col = qname, group = qname)) +
  geom_line(aes(y = y, linetype = qname), size = 1) + 
  geom_hline(data=unique(obs_data[, list(STUDY, LLOQ)]), aes(yintercept=LLOQ), linetype="dotted", size=1) +
  geom_text(data = unique(vpc$data[, list(LLOQ), by = "STUDY"]), 
            aes(x = 10, y = LLOQ, label = paste("LLOQ", LLOQ, sep = "="), ), vjust = 1, hjust = 1) +
  scale_colour_manual(name = "Simulated Percentiles\nMedian (lines) 95% CI (areas)",
                      breaks = c("q0.1", "q0.5", "q0.9"), 
                      values = c("red", "blue", "red"), 
                      labels = c("10%", "50%", "90%")) + 
  scale_fill_manual(name = "Simulated Percentiles\nMedian (lines) 95% CI (areas)", 
                    breaks = c("q0.1", "q0.5", "q0.9"), 
                    values = c("red", "blue", "red"), 
                    labels = c("10%", "50%", "90%")) + 
  scale_linetype_manual(name = "Observed Percentiles\nMedian (lines) 95% CI (areas)", 
                        breaks = c("q0.1", "q0.5", "q0.9"), 
                        values = c("dotted", "solid", "dashed"), 
                        labels = c("10%", "50%", "90%")) + 
  guides(fill = guide_legend(order = 2), colour = guide_legend(order = 2), linetype = guide_legend(order = 1)) + 
  theme(legend.position = "top", legend.key.width = grid::unit(1, "cm")) + 
  labs(x = "TIME", y = "Concentration") + 
  geom_point(data = vpc$obs, aes(x = x, y = y), size = 1, alpha = 0.1, show.legend = FALSE) + 
  geom_vline(data = bininfo(vpc)[, list(x = sort(unique(c(xleft, xright)))), by = names(vpc$strat)],aes(xintercept = x), size = rel(0.5), col = "gray80") + 
  theme(panel.grid = element_blank()) + 
  geom_rug(data = bininfo(vpc)[, list(x = sort(unique(c(xleft, xright)))), by = names(vpc$strat)],aes(x = x), sides = "t", size = 1)

```


## Plot Below Quantification Limit (BQL) 

If using the `censoring()` function, the resulting `tidyvpcobj` will also contain a pctblq table. Use `ggplot2` to plot the percentage of data below the limit of quantification across bins. 

We can include `geom_ribbon()` using the `lo` and `hi` columns in the `vpc$pctblq` table to denote the lower/upper bounds of our confidence interval. Let's also plot the median %blq of the simulated data using the `md` column in the `vpc$pctblq` table.

```{r fig.width = 9, fig.height = 6, out.width=640, warning = FALSE}
obs_data$LLOQ <- ifelse(obs_data$STUDY == "Study A", 50, 25)

vpc <- observed(obs_data, x = TIME, y = DV) %>%
  simulated(sim_data, y = DV) %>%
  censoring(blq = DV < LLOQ, lloq = LLOQ) %>%
  stratify(~STUDY) %>%
  binning(bin = NTIME) %>%
  vpcstats(qpred = c(0.1, 0.5, 0.9))

ggplot(vpc$pctblq) + 
  facet_grid(~STUDY) +
  geom_ribbon(aes(x = xbin, ymin= lo, ymax = hi), fill = "red", alpha = .2) + 
  geom_line(aes(x = xbin, y = y)) + 
  geom_line(aes(x = xbin, y = md), color = "red") + 
  labs(x= "TIME", y= "% BLQ")
```